Dissection: verified and terminating

### 2018-05-18 paper/main.tex
Edits to section 1.

\newcommand{\nonterm}[1]{\hspace*{-0.1cm}\colorbox{orange!25}{#1}}
\newcommand{\hole}[1]{\colorbox{yellow!50}{\ensuremath{\bigbox_{#1}}}}
%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'17]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}
\acmYear{2017}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

@@

\title{Verified and terminating dissections}

@@

\begin{abstract}
% The functional programming paradigm encourages a style of programming based on
%   the use of higher-order functions over inductively defined datatypes. A fold is
%   the prototypical example of such a function. However, its use for computation
%   comes at a price.  Folds for branching datatypes, such as binary trees, are by
%   definition not tail recursive functions.

%   McBride has proposed a method called
%   \emph{dissection}\cite{McBride:2008:CLM:1328438.1328474}, to
%   transform a fold into its tail-recursive counterpart. Nevertheless,
%   it is not clear why the resulting function terminates, nor it is
%   clear that the transformation preserves the fold's semantics. In
%   this paper we fill the gap by providing a fully machine checked
%   proof of the construction using the proof assistant Agda.
  \todo{Write abstract}
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{datatype generic programming, catamorphisms, dissection,
  dependent types, Agda, \todo{Keywords?}}

@@

The type of binary trees is one of the most simple, yet widespread used,
datastructures in functional programming.  Beyond its elegance and simplicity,
in its definition lies an embarasing truth: a fold over a binary tree is not a
tail recursive function. In order to understand the problem, let us introduce
the type of binary trees.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\AD{Expr}\;\mathbin{:}\;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Val}\;{}\<[10]%
\>[10]{}\mathbin{:}\;{}\<[13]%
\>[13]{}\AD{\ensuremath{\mathbb{N}}}\;{}\<[19]%
\>[19]{}\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Add}\;{}\<[10]%
\>[10]{}\mathbin{:}\;{}\<[13]%
\>[13]{}\AD{Expr}\;{}\<[19]%
\>[19]{}\to \;\AD{Expr}\;\to \;\AD{Expr}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We can write an evaluation function that maps binary trees to natural numbers if
we were to interpret the constructor \ensuremath{\AI{Add}} as addition.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{eval}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{eval}\;(\AI{Val}\;\AB{n})\;{}\<[21]%
\>[21]{}\mathrel{=}\;\AB{n}{}\<[E]%
\\
\>[3]{}\AF{eval}\;(\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}})\;{}\<[21]%
\>[21]{}\mathrel{=}\;\AF{eval}\;\AB{\ensuremath{e_1}}\;\AF{+}\;\AF{eval}\;\AB{\ensuremath{e_2}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The function \ensuremath{\AF{eval}} is compositional. The value of a node \ensuremath{\AI{Add}} is computed by
adding the values denoted by its subexpressions. And here lies the problem. The
operator \ensuremath{\AF{\_+\_}} needs both of its parameters to be evaluated before it can further
reduce. If the expression we want to compute over is very big this poses a
problem during runtime as the execution stack grows linearly with the size of
the input.

\todo[inline]{Maybe include something about Optimization through tco => get rid of
intermediate steps}

In order to solve the problem, we can make the execution stack explicit and
write a function that performs tail recursion over it. The idea is that we can
use a list to store both intemediate results and the subtress that still need to
be processed. We can define such a stack as follows:

@@

Two mutually recursive functions that operate over the stack and tree are load
and unload. The former traverses a tree finding the leftmost leaf while the
latter dispathes over the stack by accumulating a partial result while looking
for the next subtree.

@@

A tail recursive version of \ensuremath{\AF{eval}} can is defined by calling load with the empty
stack.

@@

The reader might have noticed that the names of unload and load are higlighted
with \nonterm{orange}. This is because Agda's termination checker flags the pair of
functions as possibly non-terminating. Indeed, the recursive calls are not
always performed over syntactically smaller arguments. If we assumed that it
terminates we still do not know if it is correct in the sense that for every
input \ensuremath{\AF{tail\text{-}rec\text{-}eval}} and \ensuremath{\AF{eval}} agree on the output.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{correct}\;\mathbin{:}\;\forall\;(\AB{e}\;\mathbin{:}\;\AD{Expr})\;\to \;\AF{eval}\;\AB{e}\;\AD{\ensuremath{\equiv}}\;\AF{tail\text{-}rec\text{-}eval}\;\AB{e}{}\<[E]%
\\
\>[3]{}\AF{correct}\;\mathrel{=}\;\hole{0}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

The main contribution of this paper is to show that we can both prove
termination and correctness of the construction in Agda. Specifically we show
how we can encode an ordering relation over stacks, show that the relation is
well founded and use it to define a tail recursive fold by structural recursion.
We discuss what are the main challenges that we have to overcome to convinde
Agda's termination checker that indeed something `goes smaller`. Moreover, our
approach allows us to get correctness almost for free.

McBride's insight is that we can calculate the type of stacks from the generic
description of a datatype by using Leibniz rules for derivatives. In later
sections of the paper, we discuss how to generalize our results to regular tree
datatypes and by doing so we show that McBride's intuition that the construction
is correct was true.

\section{Basic idea}

The functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} are marked as non terminating because they are
not defined by structural recursion over their arguments. In particular, the
stack passed as an argument to the recursive call of \ensuremath{\AF{load}} in the definition of
\ensuremath{\AF{unload}} is structurally equal in size as the input stack.

@@

\subsection{Correctness}

Indexing the \emph{Zipper} with an expression allow us to prove correcness of
the transformation easily. The expression during the fold does not change, thus
in every step of the computation the result of its evaluation remains constant.

@@

\section{Regular universe}
  + Universe interpretation generic programming
  + Fixpoint
  + Example??
\section{Dissection}
  + Dissection in agda
  + Plug
  + Zipper up Zipper down
  + Make clear the separation between recursion in the functor level and
  the fix level
  + relation on dissection?

>>>>

\newcommand{\nonterm}[1]{\hspace*{-0.1cm}\colorbox{orange!25}{#1}}
\newcommand{\hole}[1]{\colorbox{yellow!50}{\ensuremath{\bigbox_{#1}}}}
%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[TyDe'18]{ACM SIGPLAN Workshop on Type-driven
Development (TyDe)}{January 01--03, 2017}{St. Louis, MO, USA}
\acmYear{2018} \acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn} \startPage{1}

@@

\title{Dissection: verified and terminating}

@@

\begin{abstract}
  Lorem ipsum sic dolor amet.  Lorem ipsum sic dolor amet.  Lorem
  ipsum sic dolor amet.  Lorem ipsum sic dolor amet.  Lorem ipsum sic
  dolor amet.  Lorem ipsum sic dolor amet.  Lorem ipsum sic dolor
  amet.
  \todo{Write abstract}
\end{abstract}

@@

\keywords{datatype generic programming, catamorphisms, dissection,
  dependent types, Agda, well-founded recursion \todo{Keywords?}}

@@

Folds, or \emph{catamorphisms}, are a pervasive
programming pattern. A fold generalizes many simple traversals over an
algebraic data type. Functions implemented by means of a fold are
\emph{compositional} and structurally recursive. Consider, for
instance, the following expression data type, written in the
dependently typed programming language Agda\todo{citation}:

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{10}{@{}>{\hspre}l<{\hspost}@{}}%
\column{13}{@{}>{\hspre}l<{\hspost}@{}}%
\column{19}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AK{data}\;\AD{Expr}\;\mathbin{:}\;\AP{Set}\;\AK{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Val}\;{}\<[10]%
\>[10]{}\mathbin{:}\;{}\<[13]%
\>[13]{}\AD{\ensuremath{\mathbb{N}}}\;{}\<[19]%
\>[19]{}\to \;\AD{Expr}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\AI{Add}\;{}\<[10]%
\>[10]{}\mathbin{:}\;{}\<[13]%
\>[13]{}\AD{Expr}\;{}\<[19]%
\>[19]{}\to \;\AD{Expr}\;\to \;\AD{Expr}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
We can write a write a simple evaluator, mapping expressions to
natural numbers, as follows:
%
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{21}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\AF{eval}\;\mathbin{:}\;\AD{Expr}\;\to \;\AD{\ensuremath{\mathbb{N}}}{}\<[E]%
\\
\>[3]{}\AF{eval}\;(\AI{Val}\;\AB{n})\;{}\<[21]%
\>[21]{}\mathrel{=}\;\AB{n}{}\<[E]%
\\
\>[3]{}\AF{eval}\;(\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}})\;{}\<[21]%
\>[21]{}\mathrel{=}\;\AF{eval}\;\AB{\ensuremath{e_1}}\;\AF{+}\;\AF{eval}\;\AB{\ensuremath{e_2}}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
%
In the case for \ensuremath{\AI{Add}\;\AB{\ensuremath{e_1}}\;\AB{\ensuremath{e_2}}}, the \ensuremath{\AF{eval}} function makes two recursive
calls and sums their results. Such a function can be implemented by a
fold, passing the addition and identity functions as the argument
algebra.

Unfortunately, all in the garden is not rosy. The operator \ensuremath{\AF{\_+\_}}
needs both of its parameters to be fully evaluated before it can
reduce further. When evaluating large expressions, the size of the
stack used during execution grows linearly with the size of the input,
potentially leading to a stack overflow, causing the execution of the
program to halt unexpectedly.

To address this problem, we can rewrite the evaluator to be
\emph{tail recursive}. Modern compilers typically map tail recursive
functions to machine code that runs in constant memory. To write such
a tail recursive function, we need to introduce an explicit stack
storing both intemediate results and the subtrees that have not yet
been evaluated. We can define such a stack as follows:

@@

We can define a tail-recursive evaluation function by means of a
pair of mutually recursive functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}}. The \ensuremath{\AF{load}}
function traverses the input expression, pushing subtrees on the
stack; the \ensuremath{\AF{unload}} function unloads the stack, while accumulating a
(partial) result.

@@

We can now define a tail recursive version of \ensuremath{\AF{eval}} by
calling \ensuremath{\AF{load}} with an initially empty stack:

@@

Implementing this tail recursive evaluator has come at a price: Agda's
termination checker flags the \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} functions as
potentially non-terminating by highlighting them
\nonterm{orange}. Indeed, in the very last clause of the \ensuremath{\AF{unload}}
function a recursive call is made to arguments that are not
syntactically smaller. Furthermore, it is not clear at all that the
tail recursive evaluator produces the same result as our original
one. It is precisely these questions that this paper tackles by
means of the following novel contributions:

\begin{itemize}
\item We give a verified proof of termination of \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
  a carefully chosen \emph{well-founded relation}
  (Section~\ref{sec:basics}). After redefining \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
  this relation, we prove that it agrees with the \ensuremath{\AF{eval}} function on
  all inputs.
\item We generalize this construction, inspired by McBride's
  generalization of Danvy's work on abstract machines. To do so, we
  define a universe of algebraic data types and a generic fold
  (Section~\ref{sec:universe}). Subsequently we show how to turn any
  structurally recursive function defined using a fold into a tail
  recursive variant (Section~\ref{sec:dissection}). 
\item Finally, we sketch how the proofs of termination and semantics
  preservation from our example are generalized to the generic fold
  over arbitrary types in our universe (Section~\ref{correctness}).
\end{itemize}

All the constructions and proofs presented in this paper have been
implemented in Agda\todo{citation}. The corresponding code is freely
available online.\footnote{\todo{url}}

\section{Basic idea}
\label{sec:basics}
The functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} are marked as non terminating
because they are not defined by structural recursion over their
arguments. In particular, the stack passed as an argument to the
recursive call of \ensuremath{\AF{load}} in the definition of \ensuremath{\AF{unload}} is structurally
equal in size as the input stack.

@@

\subsection{Correctness}
\label{sec:basic-correctness}
Indexing the \emph{Zipper} with an expression allow us to prove
correcness of the transformation easily. The expression during the
fold does not change, thus in every step of the computation the result
of its evaluation remains constant.

@@

\section{Regular universe}
\label{sec:universe}
  + Universe interpretation generic programming
  + Fixpoint
  + Example??
\section{Dissection}
\label{sec:dissection}  
  + Dissection in agda
  + Plug
  + Zipper up Zipper down
  + Make clear the separation between recursion in the functor level and
  the fix level
  + relation on dissection?

\section{Termination and correctness, generically}
\label{sec:dissection}
